Peer Assignment #1
========================================================

### Loading and preprocessing the data
```{r Read Data}
setwd("C:/Users/hwildasi/Desktop/Coursera")
active <- read.csv("activity.csv")
```

### What is the mean total number of steps taken per day?
```{r Calculate Mean Per Day}
day <- tapply(active$steps, active$date, sum)
day_mean <- mean(day, na.rm=TRUE)
day_median <- median(day, na.rm=TRUE)
hist(day,
     main="Histogram of Total Number of Daily Steps",
     xlab="Total Number of Daily Steps")
```

The mean number of steps per day is `r day_mean` and the median number of steps per day is `r day_median`.

### What is the average daily activity pattern?

```{r Time Serids}
steps_int <- aggregate(steps~interval, active, mean, na.rm=TRUE)
plot(steps_int$interval, steps_int$steps, type="l",
     main="Plot of Mean Steps per 5-minute Interval", xlab="5-minute Interval",
     ylab="Mean Number of Steps")
max <- steps_int[which.max(steps_int$steps),1]
```

The `r max` 5-minute interval contains the maximum number of steps on average across all the days.

### Inputing missing values

```{r Missing Data}
steps <- active$steps     #Create a vector of steps
miss <- is.na(steps)      #Find missing/non-missing
nmiss <- steps[miss]      #Only keep missing values
num_miss <- length(nmiss) #Count number of elements in vector
```

A total of `r num_miss` values are missing in the dataset.

For records with a missing value for the number of steps. Impute the value as the average for that time interval across all days and compute corresponding histogram.
```{r Imput Missing values}
names(steps_int)[2] <- "steps_avg"  #rename steps to steps_avg
active_imp <- merge(active, steps_int, by="interval")
active_imp$stepsi <- max(active_imp$steps, active_imp$steps_avg)
active_imp$stepsi <- ifelse(is.na(active_imp$steps), active_imp$steps_avg, active_imp$steps)
```

```{r Histogram with imputed data}
dayi <- tapply(active_imp$stepsi, active_imp$date, sum)
dayi_mean <- mean(dayi, na.rm=TRUE)
dayi_median <- median(dayi, na.rm=TRUE)
hist(dayi,
     main="Histogram of Total Number of Daily Steps - Missing Values Imputed",
     xlab="Total Number of Daily Steps")
comp_mean <- ifelse(dayi_mean == day_mean, "equal",
                      ifelse(dayi_mean > day_mean, "greater", "lesser"))
comp_median <- ifelse(dayi_median == day_median, "equal", 
                      ifelse(dayi_median > day_median, "greater", "lesser"))
diff_mean <- abs(day_mean - dayi_mean)
diff_median <- abs(day_median - dayi_median)
diff_sum <- ifelse(diff_mean < 10 & diff_median < 10, "insignificant (<10 steps per day)",
                    ifelse(diff_mean<500 & diff_median<500, "minimal (<500 steps per day)", 
                           "Significant (>500 steps per day)"))
```

The mean number of steps per day using imputed values is `r dayi_mean` and the median number of steps per day is `r dayi_median`. In comparison to the values generated excluding missing values the mean is `r comp_mean` and the median is `r comp_median`. The change in the mean and median values is `r diff_sum`.

### Are there differences in activity patterns between weekdays and weekends?
```{r Weekdays}
#Create Factor Levels for Weekday and Weekend
active_imp$daten <- as.Date(active_imp$date, "%Y-%m-%d")
active_imp$weekday <- weekdays(active_imp$daten)
active_imp$weekdayf <- ifelse(active_imp$weekday=="Saturday" | active_imp$weekday=="Sunday",
        "weekend", "weekday")

#Sum across weekday/weekend and plot data
stepsi_int <- aggregate(stepsi~interval*weekdayf, active_imp, mean, na.rm=TRUE)
stepsi_int$interval <- as.numeric(stepsi_int$interval)
library(lattice)
xyplot(stepsi_int$stepsi~stepsi_int$interval | stepsi_int$weekdayf,type="l",
     main="Plot of Mean Steps per 5-minute Interval - Weekend vs. Weekday", xlab="5-minute Interval",
     ylab="Mean Number of Steps", layout=c(1, 2))
```

